Two Problems for Sophistication

### 2015-01-21 Facticity/Facticity/paper.tex
discussion

\section{Discussion}
\label{section:conclusion}

We have criticized existing measures of the sophistication of a binary string. Our results show that there are technical problems with each of them. But that does not in itself mean that it should be impossible to come up with a sound measure. The intuition throughout the literature, starting with Kolmogorov's proposal of the structure function, appears to have been that it should be possible to identify noise, and distinguish it from structural information, by its property of being incompressible.

@@

For the first argument, we take a generative perspective. If data are generated by a computable probability distribution, then a sophistication could be called \emph{consistent} if, given a large enough sample, it accurately reflects the complexity of that distribution. While in general we do not wish to make any assumptions about the origin of the data, it can make sense to do so as a sanity check, as we will do here. Suppose that the data are generated by a universal distribution, e.g. by feeding random bits to a universal partial computable function. Let $u$ be the index of the function in the $\phi$-enumeration with the property that $\phi_u(\bar\imath y)=\phi_i(y)$. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of data that is generated in this fashion?

@@

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. There are many good models for this string, ranging from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal models says that it is `something computable'. A more specific model might say that it is `an image'. Even more specific would be a painting, a Monet, or specifically the painting Impression of a Sunrise. A sound measure of Sophistication should be able to select one of these models as the proper representation of structure in the data, and disqualify the others as either over- or underfitting. But how should we be able to say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? Intuitively such distinctions require further assumptions.

The second argument why we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Figure~\ref{fig:diagram} provides an illustration. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in the figure. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. \emph{However}, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps do not occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that \emph{part} of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=\min\{|y|:\phi_i(y)=x\}-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data --- if $\phi_i$ truly generated the data, then the probability that the random deficiency is more than $k$ bits is less than $2^{-k}$ (proof omitted). But fundamental problems arise as soon as hard cut-offs are introduced. We also wish to point out that a lot of measures taken in the literature to address problems with sophistication, such as restricting the model class or introducing model penalties, are no longer necessary as soon as such hard cut-offs are avoided. What remains is a simple and elegant theory that can disprove most models, leaving not one, but a limited number of interesting candidates.

\section{Conclusion}
As we have seen, the idea behind sophistication has a rich history. It has been proposed many times, by many different authors. This tells us two things. First, there is a strong intuition behind sophistication, shared by many respected authors, and second, it is not easy to get the definition exactly right.
 While no authors have touched on the issue of invariance, it is related to the issue of instability \cite{antunes2013sophistication,vereshchagin2013algorithmic}. We consider invariance a first criterion that any proposal must satisfy, , with the burden of proof resting with the proposer, rather than the critic.

Finally, we note that the use of total functions has a very subtle effect on the discussion surrounding sophistication. Since the total functions are not enumerable, the value of the sophistication relies strongly on the difference between the Kolmogorov complexity and the Kolmogorov complexity restricted to total functions. This gap is exceedingly difficult to capture, because the total functions cannot be enumerated. This is dangerous: if such definitions cannot be proved wrong or right, they will stand as the ad-hoc official view, while the restrictions used to elude proofs of incorrectness make them very difficult to build on. This creates an artificial dead end for a very valuable area of research.

>>>>

\section{Discussion and conclusion}
\label{section:conclusion}

We have criticized existing measures of the sophistication of a binary string. Our results show that there are technical problems with each of them. But that does not in itself mean that it should be impossible to come up with a sound measure. The intuition throughout the literature, starting with Kolmogorov's proposal of the structure function, appears to be that it should be possible to identify noise, and distinguish it from structural information, by its property of being incompressible.

@@

For the first argument, we take a generative perspective. If data are generated by a computable probability distribution, then a sophistication could be called \emph{consistent} if, given a large enough sample, it accurately reflects the complexity of that distribution. Suppose that the data are generated by a universal distribution, e.g. by feeding random bits to a universal partial computable function. (While in general we do not wish to make any assumptions about the origin of the data, it can make sense to do so as a sanity check, as we will do here.) Let $u$ be the index of the function in the $\phi$-enumeration with the property that $\phi_u(\bar\imath y)=\phi_i(y)$. Then the initial random bits will determine the prefix encoded index $\bar\imath$ of the function $\phi_i$ that $\phi_u$ will subsequently emulate, and the remaining bits are used as inputs to $\phi_i$. We now ask, what should be the sophistication of data that is generated in this fashion?

@@

Consider the following metaphor. We are given a bitstring which can be read as a bitmap image of the painting \emph{Impression of a Sunrise}. There are many good models for this string, ranging from very generic to very specific. The theory behind sophistication suggests that we can choose one of these as the objective, intrinsic model of the data. Different models say different things: the universal model says that it is `something computable'. A more specific model might say that it is `an image'. Even more specific would be a painting, a Monet, or specifically the painting Impression of a Sunrise. A sound measure of Sophistication should be able to select one of these models as the proper representation of structure in the data, and disqualify the others as either over- or underfitting. But how should we be able to say that the data is intrinsically more of an image than a painting? More of a Monet than a painting? Intuitively such distinctions require further assumptions.

The second argument why we believe sophistication will elude definition is more technical. Given an input $x$ and some model class, say $\C$, consider the set of all possible two-part representations of $x$. Figure~\ref{fig:diagram} provides an illustration. Now, when the numbering of partial computable functions is changed, the codelength of the model part of all these representations will change. This is illustrated in the right graph in the figure. However, the Invariance Theorem expresses that this change is limited by a constant term that does not depend on $x$. So for sufficiently large $x$, the horizontal shift of all the representations will be relatively small. \emph{However}, even this small shift can push some representations out of the acceptable region (indicated by the dashed line), and pull others in. This may lead to a different representation determining the sophistication, one whose \emph{total} codelength is close to what it was before, but whose \emph{model} codelength can be anywhere between $0$ and $C^\C(x)$. If such jumps can occur, the sophistication is not invariant. And while we cannot \emph{prove} in general that such jumps can alwys occur, there seems to be no reason to believe that they don't.

For this reason, we take a skeptical view of sophistication. Note that \emph{part} of the theory is fine: there is nothing wrong with evaluating models for the data by comparing their two-part code lengths. In fact, the randomness deficiency $\delta(\phi_i,x)=\min\{|y|:\phi_i(y)=x\}-C^\K(x\mid i)$ of a model $\phi_i$ has a direct statistical interpretation as a measure of counterevidence against the hypothesis that it generated the data --- if $\phi_i$ truly generated the data, then the probability that the random deficiency is more than $k$ bits is less than $2^{-k}$ (proof omitted). In the Monet example above, this will allow us to disqualify the model expressing that the data is actually, say, a recording of Jazz music.

But fundamental problems arise as soon as hard cut-offs are introduced. In our opinion, a lot of measures taken in the literature to address problems with sophistication, such as restricting the model class or introducing model penalties, complicate the method and make problems harder to analyse, without actually addressing the fundamental issue. This is dangerous: if such ad-hoc ``fixes'' result in a theory that is hard to prove either wrong or right, then this can create an artificial dead end for a valuable area of research. When hard cut-offs are avoided on the other hand, all such measures are suddenly no longer necessary. What remains is a simple and elegant theory that can be used to sift through all possible models, disproving most while retaining a select number of interesting candidates for our further consideration.
